Reconstructing the last known movements of one of Nature's giants

### 2017-11-07 manuscript/revision/response.md

Dear Editors,

Our paper was reviewed at Nature and was not accepted for publication. However, it was suggested that we transfer the paper to Nature Ecology and Evolution, which we are glad to do. We are transferring the reviews with the paper, therefore we include a response to referees as part of this cover letter. We are grateful to the referees for their helpful and constructive comments, and hope that our edits, along with the move from Nature to Nature Ecology and Evolution, will have allayed their concerns about the manuscript.

We respond to their points below. To improve the clarity of this document we have included our responses to the referees comments in italics.

**Important note to the referees**. During the revisions of this manuscript, which included substantial extra analysis, we discovered an error in our original results. Essentially, the baleen samples had been labeled in the reverse direction to what we expected. The values for all the stable isotope measures remain the same but the timeline is reversed (this occurred due to a misunderstanding with the masters student who ran to original mass spec analyses and subsequently finished her masters project and moved away). We regret this error, but these things happen, and we are relieved that we noticed the error before publication. Interestingly, when we added the extra data to the end of our time series, we find our overall results do not change from our previous manuscript. We communicated with Editor in Chief Patrick Goymer about whether it was still appropriate to cascade our reviews with the paper given this error, and he said this was fine as long as we made this very clear in our response letter. This note is to alert the referees that the results graphs will look different to the original version, though our conclusions have not changed qualitatively.

## Original referee comments

@@

In my opinion, the paper makes two important points:

First: The authors used a mix between a temporal isoscape (measured POM over time and space to create a changing isoscape) and an agent-based model (AI that models the movement of an individual using parameters such as presence of food, time and temperature in adjoining locations to decide if it moves or not). The combination was used to retrace the movement of a blue whale and compare it to baleen isotope records. In my opinion, this is a novel approach that provides more confidence in attributing isotopic signals to place/time, and thus helps reconstruct migration of the individual. However, I feel that it is not a solve-all approach, as just as with isoscapes, you still need to get those initial measurements of time/space isotopes. Nevertheless, it is a very interesting technique.

Second: the baleen isotope records themselves are interesting as there are very few blue whale baleen records published (this is the second one ever to my knowledge, and the first was also published this year). The isotopes also appear to show a mixture of behaviours, such as migration AND permanent residence in a single individual. Such mix of behaviour is rare in the literature, and still debated. Further, the residence appears consistent with pregnancy/lactation, which is even more uncommon in whale baleen literature as far as I know. For these reasons, I think the isotope records will be very interesting to other researchers in the field.

> We thank the referee for their positive comments about our paper.

@@

### Data and Methodology: Validity

I am satisfied with the SIA methods and results, but I have some slight comments. While I have no experience with modelling, I noted that there was no mention of isotopic fractionation in the model. As far as I know, blue whale feed on zooplankton and small fish, not directly on phytoplankton. If I am correct, the comparison of the baleen records to the modelled isotopes may be incorrect, as it should be affected by trophic shifts. I suppose it depends on whether the model was used for a direct pattern comparison, or if the values modelled and the baleen values were compared to each other.

> The referee is correct

* CLIVE how are we dealing with this???

Comparable literature in other species (seals, humans) appear to confirm the pregnancy interpretation, although I'm less convinced by the lactation signal as it seems too delayed in time. This could possibly be checked using hormones, although that is also a very novel field. See the following literature on baleen hormones: Hunt 2016 “Longitudinal progesterone profiles in baleen from female North Atlantic right whales (Eubalaena glacialis) match known calving history”; Hunt 2014 “Baleen hormones: a novel tool for retrospective assessment of stress and reproduction in bowhead whales (Balaena mysticetus)”. These are older samples (20 yrs or so) but not as old as this whale so may be loss of hormone signal over time.
I strongly believe this will not prohibit publication, as the results are still very interesting. It is very possible that I am misreading the delay in lactation signal.

> Measuring hormone levels in the baleen would be a great addition to this work. We have colleagues in the US who are planning to extract hormones from this baleen. Unfortunately these results have been delayed now for over a year due to US permitting issues (thanks Trump!) and the CITES status of blue whales. But we agree that this would be really interesting, if the hormone signals remain over time.

* LACTATION SIGNAL - comments?

@@

There are no error bars or replication of samples, but this is typical of stable isotopes.

* Do we need a comment on this?

### Suggested improvements/ References:

Some basic stable isotope knowledge, as well as blue whale feeding ecology should probably be mentioned (ie krill/fish feeding, no fasting) or trophic influence on isotopic signal compared to model POM values. Adding sentence on feeding ecology at beginning, and explaining in model why POM is a good proxy for zooplankton should be enough. For example: Line 65: a reference such as West et al, 2006 “Stable isotopes as one of Nature’s ecological records” relevant to SIA basics. Line 82: reference again with basic SIA shifts in marine food webs? Possibly McMahon et al, 2015 “Millennial Scale plankton regime shifts in the subtropical North Pacific Ocean”.

> We have added this as suggested

>

* Add it!

The reported baleen growth rates fit the literature, but I believe it may be worth mentioning in Line 61 that baleen gets worn down over time, and that the records are not full life spans.

> We have clarified this in the text at line XXX as follows: "It is ideal for stable isotope studies because keratin grows continuously through an individual's life (although it is worn away at the tips over time, so rarely records an individual's entire lifespan), and once laid down it is metabolically inert."

According to the model, d15N is relatively higher in the arctic during summer than the Atlantic in winter. In the profile, the winter values in phase one are still comparatively higher than the summer values. Would you be able to explain? I am mostly curious, but it may reflect difficulties with reading the model from a figure.

> Explain

While the d13C increase fits the sea lion theory, the drop in d13C seen after/at birth happens too quick (late 1889) in the whale to fit the d15N rise consistent with lactation (Almost 6 months late according to timeline, past mid 1890). Human analyses in hair showed a decrease in d15N (1%o) during pregnancy, which can be seen in the baleen if we take the d13C drop as birth time and early 1889 as pregnancy time (see Fuller et al 2004 “Nitrogen Balance and 15N, why you're not what you eat during pregnancy”). While it's not a study conducted in marine mammals, it's a further indication that the birth time may indeed be in late 1889, or at least that the lactation peak is very late. Could indicate a slowing down in baleen growth, or something else? Could be worth checking with hormonal analyses (see above).

> Explain

Another possibility lies in the location of baleen analyses. Since the methods are not clear on that point, can you confirm that the samples were taken past the gum line? Dried skin/gum could have affected the isotopic signals if samples were taken outside of the erupted zone. Could it be why the isotopic signals are so different in phase two? Busquet Vass paper shows no difference in 13C between skin and baleen so unlikely, but worth confirming.

> Check

The final sample interpretation of a return to cyclicity is very open to discussion, as a single point is too weak to be a guarantee. I will concede that the isotopes are consistent with past migratory values, but may be worth softening the comment (line 109-110)?

> OK

When describing the different behaviours, such as whales staying in high latitudes instead of migrating, also talk about the ones staying at low lat (isotope blue whale ref. Busquet-Vass) since it is more relevant to your own isotopic interpretation (line 127).

> Alright

@@

Typos etc:

1. Abstract
Line 20: Should not have abbreviation. Move to line 58 where it is introduced next.

> Fixed

2. Article main text
Line 56: “)” extraneous bracket

> Fixed

3. Supplementary Data
Page 2, line 1: “)” extraneous bracket after “behavioural phase 2”

> Fixed

@@

### Clarity and Context:

The manuscript was clear, but lacked some information on basic isotope literature, and blue whale ecology. While the ecology side is less important, some information on how isotope analyses work could be beneficial to the greater public. I understand that the limited space available prevents a thorough explanation of SIA, but there is no explanation of the general differences between d13C and d15N (geographical versus trophic information). I believe a quick sentence describing the two tracers in line 57 (for example) could help.

> We have added a brief comment on what blue whales eat on line XXX:.

> We have also added some simpler references to stable isotopes. For example. We also calrify the units used as this often trips up new users: lin XXX "Note that carbon and nitrogen isotopes are expressed as $\delta^{13}$C and $\delta^{15}$N values, i.e. relative to a standard and as ratios of the rare ($^{13}$C, $^{15}$N) to common ($^{12}$C, $^{14}$N) isotope (West et al 2006)."
"

The abstract was clear and direct, providing a good summary of the paper's approach and conclusions.
The conclusion is clear and justified, although the claim of first quantitative evidence of movement behaviours in blue whale seems unlikely given the slightly older paper from Busquets-Vass also reporting blue whale baleen isotopes.

> Clarify

## Referee 2:

> If we are interpreting Referee 2's comments correctly (see below), we think they mostly revolve around issues about the novelty of the study, and overstatement of results based on one individual. We completely see the points the referee is making. Some of this was related to the difficulties of writing in the shorthand Nature style - we have clearly been a little too casual in some of our wording. In other cases we haven't been clear that the selling point of this method is that it can easily be applied to hundreds of individuals at relatively low cost. Here we test the methods on one individual, but the real value in the method is its broader applicability. We have tried to clarify this where possible and have also toned down/removed/clarified a lot of our previous statements about the novelty of the study and interpretations of the results. We explain our changes in more detail below.

line XXX: These methods can be applied to large numbers of individuals, but as proof of concept, here we apply this method to baleen from the Natural History Museum (NHM) London's blue whale `Hope'.

I appreciated the opportunity to review the submitted manuscript “Reconstructing the last known movements of one of Nature's giants”.

@@

140-141: OK...but it is movement behavior in a single blue whale, and this has been done in larger sample sets of other baleen whales.

>

147-149. This sentence makes it seem like hardly anyone is using museum specimens for SIA or other chemical analyses, which simply isn't the case.

>

> We hope we have responded to all these comments appropriately. Please let us know if you require any further information,

> Clive Trueman, on behalf of the authors.

### 2017-11-07 manuscript/revision/blue-whale-ms-v2.tex
Edits to manuscript and cover letter

\raggedright

\paragraph{Understanding animal movements is crucial for effective conservation, especially as species ranges shift due to climate change\cite{runge2014conserving,robinson2009travelling}. 
Animal tissues housed in museum collections provide biochemical records of behaviour that can be used to reconstruct individual-level movements over long timescales and under historical climatic and ecological conditions\cite{newsome2010using}. 
Here we combine stable isotope analyses of baleen with novel agent-based simulation models, to reconstruct the movements of an iconic blue whale (\textit{Balaenoptera musculus}) that stranded off Ireland in 1891 and is now on display at the Natural History Museum, London (NHM). 
We identify two distinct phases in the whale's behaviour: repeated annual north-south migrations in sub-Arctic waters, and residency in subtropical waters. 
Subtropical residency is associated with isotopic signatures consistent with pregnancy and lactation in the year before the whale's death. 
These results supply the first quantitative multi-year records of historical migrations in individual blue whales, and suggest that blue whale residency and movement patterns in the North Atlantic have remained similar for hundreds of years, despite changing human pressures in the region. 
Historic tissue collections can provide unique insight into individual-level movement behaviours associated with changing climate, ecological conditions and anthropogenic pressures, and reveal aspects of life history. 
Combining biochemical tracers with simulation modeling unlocks the information recorded in these irreplaceable archives.}

@@

The blue whale (\textit{Balaenoptera musculus}), the largest animal to have ever lived, is one such migratory marine species. 
Despite blue whale hunting being banned for over 50 years, blue whales are still listed as Endangered by the IUCN Red List\cite{reilly2008balaenoptera}.
Conservation efforts are hampered by the scarcity of detailed data on the distribution, behaviour and ecology of blue whales, both for historical and present-day populations. 
Current methods for tracking large whales include satellite tracking, radio tagging, aerial photography, passive acoustic monitoring, and sightings records\cite{borger15,mcdonald2006biogeographic,bailey2009behavioural,mate2007evolution,silva2013north}. 
These methods are adding vital behavioural data to our understanding of whale movements, however, they are often expensive, limited to short-term studies on few individuals\cite{bailey2009behavioural,best2015tag,mate2007evolution}, and may influence whale behaviour\cite{walker2012review}. 
Intrinsic biogeochemical tags provide an alternative technique, where the chemical compositions of animal tissues are compared to spatially referenced models of chemical variation across the species' range\cite{hobson1999tracing,hobson2008tracking}). 
Stable isotope-based geolocation has been used to interpret whale ecology \cite{hobson1998stable,ryan2013stable,eisenmann2016isotopic}. 
Here we apply this method to baleen from the NHM's blue whale.

@@

Baleen is formed of plates of keratin and is used by mysticete whales to filter feed. 
It is ideal for stable isotope studies because keratin grows continuously through an individual's life, and once laid down it is metabolically inert\cite{best1996stable}.

%Refs to support long history of inferring movement from baleen isotopes.

We sampled keratin from the blue whale's baleen plate at 1cm intervals, perpendicular to the direction of growth, and used continuous-flow isotope ratio mass spectrometry to determine the isotopic compositions of carbon and nitrogen (expressed as $\delta^{13}$C and $\delta^{15}$N values). 
Given the date of stranding (25th March 1891), and estimated baleen growth rates of 13.5cm $y^{-1}$ (see Supplementary Methods), we reconstructed a timeline for $\delta^{13}$C and $\delta^{15}$N fluctuations in the baleen over the seven full years of the whale's life (early 1884 - spring 1891). 
The $\delta^{13}$C and $\delta^{15}$N profiles show two distinct phases in the whale's behaviour. 
Throughout the record, $\delta^{15}$N values record regular cyclical fluctuations, with a mean spacing of 13.5cm, assumed to represent annual periodicity (see Supplementary Methods).  
In behavioural phase one (from record start in early 1884 to spring 1886), we find relatively stable, elevated $\delta^{13}$C values, and relatively low $\delta^{15}$N values (Fig. \ref{fig1}; Figs. S1 and S2). 
In behavioural phase two (summer 1886 to spring 1891) $\delta^{15}$N values are relatively elevated and we see coincident cyclical fluctuations in both $\delta^{13}$C and $\delta^{15}$N values. 
In the last year of life the cyclical pattern is disrupted, with constant low $\delta^{13}$C values for approximately six months in the first half of 1890, before a rapid switch to relatively positive $\delta^{13}$C values in the second half of 1891. 
The final three months of the record show a continued fall in $\delta^{13}$C values (Fig. \ref{fig1}; Figs. S1 and S2).

@@

% figure 1
\begin{figure}
  \centering
  \includegraphics[width = \linewidth]{figures/Figure-1-raw-dC-dN-data.png}
  \caption{Variation in $\delta^{13}$C (black circles, left y-axis) and $\delta^{15}$N (grey circles, right y-axis) longitudinally through the baleen plate (n = 77 samples from a single baleen plate for both isotopes) of the NHM blue whale exhibits strong annual periodicity and cross-correlation (Figs. S1 and S2) in both isotopes as well as a change in the mean and magnitude of the cycles during 1889 (approximately at sample 30cm). Approximate relationship to years assuming a growth rate of 13.5 cm y$^{-1}$ is shown on the upper x-axis, and year boundaries are indicated by vertical dotted grey lines.}
  \label{fig1}
\end{figure}

Relating marine animal tissue isotopic compositions to likely locations is complicated by both a lack of spatial reference values, and because the isotopic composition of phytoplankton at the base of marine food webs varies in both space and time. 
We approach these problems using newly developed, temporally dynamic models of $\delta^{13}$C values in the global ocean\cite{magozzi2017using} (Fig. S3), coupled to an agent-based model of whale movements (see Supplementary Methods; Table S1; Fig. S5) to simulate time series of $\delta^{13}$C values expected under differing movement behaviours and in differing parts of the known geographic range of blue whales. 
We then compare simulated profiles (Figs. \ref{fig2}, \ref{fig3} and Fig. S4) with the measured isotopic profile of the blue whale's baleen (Fig. \ref{fig1}) to identify the most likely movement history.
We assume continued ingestion of new protein throughout the year. 
Other mysticete whales fast during migrations and during winter, relying on stored lipid reserves to meet metabolic demands, but blue whales are too large to stop feeding for long periods; \cite{busquets2017estimating,silva2013north}).
%Can we back this up? Are we more saying that we think they need protein to maintain tissue turnover but could meet energetic demands form stored lipids?

@@

The last 500 days of the whale's life are difficult to simulate. 
Beginning in the winter of 1889/1890, the observed $\delta^{13}$C profile is relatively depleted, and remains constant for approximately six months, before increasing rapidly in the winter of 1890/1891. 
Depleted $\delta^{13}$C values are found in northern waters that show large temporal fluctuations in $\delta^{13}$Cplk values [fig SX residency plot], and the observed values cannot be simulated purely from movements within the known geographic range.
We infer that the $\delta^{13}$C record in the last period of life reflects the physiological demands of pregnancy, as similar increases in $\delta^{13}$C values have been seen in pregnant South American sea lions\cite{cardona2017temporal}. 
We suggest that the constant, depleted $\delta^{13}$C values reflect use of carbon reserves assimilated in northern latitudes during 6-8 months residency in low latitude waters. 
After six months the whale began feeding in low latitude waters producing a dramatic increase in $\delta^{13}$C values and began a final migration north. 
Blue whales have a 10-12 month gestation period, with calving occurring in sub-tropical waters, and calves are weaned after 6-7 months\cite{handbook}. 
The proposed timescale of pregnancy around early spring 1889, birth in spring 1890, residency in warm waters through the whole of 1890 with feeding resuming in the second half of 1890 and eventual stranding during the return to northern feeding grounds in early 1891, is therefore consistent with the known reproductive ecology of blue whales.

@@

Conventional wisdom suggests that baleen whales migrate each year from feeding grounds near the poles in summer, to overwintering and breeding grounds nearer the equator\cite{corkeron1999baleen,lockyer1981migration}. 
However, a growing body of evidence suggests that not all whales migrate to subtropical waters each year, with some remaining at high latitudes year round\cite{mcdonald2006biogeographic}. 
Our results support this conclusion. 
Interestingly, there is still a resident population of blue whales in the regions indicated by our model, and females still breed in the subtropical waters near the Azores\cite{reilly2008balaenoptera}. 
Whaling was an intense pressure for blue whales during the period we are analysing. 
Before whaling in the North Atlantic began in 1868\cite{reilly2008balaenoptera}, there were 10,000-15,000 blue whales in the region\cite{sigurjonsson1995life}. 
In the early 20th century, fisheries moved outside the area because stocks were so depleted\cite{reilly2008balaenoptera}; during this period over 12,000 blue whales were landed\cite{sigurjonsson1995life}. 
The NHM blue whale thus represents an individual from a species at the brink of extinction.
Our data suggest that the spatial ecology of blue whales in the North Atlantic has remained similar for more than 100 years, despite dramatic changes in anthropogenic pressures.

This is the first quantitative evidence of historical individual-level movement behaviours in blue whales. 
Our novel simulation modeling removes a long-standing limitation in stable isotope ecology, and can be applied to stable isotope records from any incrementally-grown tissue to estimate most likely individual movement behaviours over multiple years. 
Such techniques can also be applied to extinct species behaviours and used to understand how populations were affected by past pressures such as hunting, and present-day pressures such as global change. 
Finally, our results highlight that museums represent a vast, mostly untapped, source of individual-level spatial ecology information.

>>>>

\raggedright
% Need to edit
\paragraph{Understanding animal movements is crucial for effective conservation, especially as species ranges shift due to climate change\cite{runge2014conserving,robinson2009travelling}. 
Animal tissues housed in museum collections provide biochemical records of behaviour that can be used to reconstruct individual-level movements over long timescales and under historical climatic and ecological conditions\cite{newsome2010using}. 
Here we combine stable isotope analyses of baleen with novel agent-based simulation models, to reconstruct the movements of an iconic blue whale (\textit{Balaenoptera musculus}) that stranded off Ireland in 1891 and is now on display at the Natural History Museum, London. 
We identify two distinct phases in the whale's behaviour: repeated annual north-south migrations in sub-Arctic waters, and residency in subtropical waters. 
Subtropical residency is associated with isotopic signatures consistent with pregnancy and lactation in the year before the whale's death. 
These results supply the first quantitative multi-year records of historical migrations in individual blue whales, and suggest that blue whale residency and movement patterns in the North Atlantic have remained similar for hundreds of years, despite changing human pressures in the region. 
Historic tissue collections can provide unique insight into individual-level movement behaviours associated with changing climate, ecological conditions and anthropogenic pressures, and reveal aspects of life history. 
Combining biochemical tracers with simulation modeling unlocks the information recorded in these irreplaceable archives.}

@@

The blue whale (\textit{Balaenoptera musculus}), the largest animal to have ever lived, is one such migratory marine species. 
Despite blue whale hunting being banned for over 50 years, blue whales are still listed as Endangered by the IUCN Red List\cite{reilly2008balaenoptera}.
Conservation efforts are hampered by the scarcity of detailed data on the distribution, behaviour and ecology of blue whales, both for historical and present-day populations. 
Current methods for tracking large whales include satellite tracking, radio tagging, aerial photography, passive acoustic monitoring, and sightings records\cite{borger15,mcdonald2006biogeographic,bailey2009behavioural,mate2007evolution,silva2013north}. 
These methods are adding vital behavioural data to our understanding of whale movements, however, they are often expensive, limited to short-term studies on few individuals\cite{bailey2009behavioural,best2015tag,mate2007evolution}, and may influence whale behaviour\cite{walker2012review}. 
Stable isotope analyses provide an alternative technique\cite{west2006stable}. 
We can use the isotopic compositions of whale tissues to reconstruct their movement patterns, by comparing them to spatially referenced models of isotopic variation across the species' range\cite{hobson1999tracing,hobson2008tracking,eisenmann2016isotopic}. 
%line 57,
These methods can be applied to large numbers of individuals, but as proof of concept, here we apply this method to baleen from the Natural History Museum (NHM) London's blue whale `Hope'.

@@

Baleen is formed of plates of keratin and is used by mysticete whales to filter feed. 
It is ideal for stable isotope studies because keratin grows continuously through an individual's life (although it is worn away at the tips over time, so rarely records an individual's entire lifespan), and once laid down it is metabolically inert\cite{best1996stable}. 
Blue whales feed mainly on krill (Euphausiacea), small crustaceans that feed on phytoplankton.
Thus the stable isotope compositions of keratin at a given point in the baleen will reflect the stable isotope compositions of the krill it was feeding on around two weeks prior to that time point (two weeks is required for the carbon and nitrogen to be assimilated and converted into tissues). %REF???
Baleen stable isotopes have long been used in this way /.... % finish
%Refs to support long history of inferring movement from baleen isotopes.

We sampled keratin from the blue whale's baleen plate at 1cm intervals for 100cm, perpendicular to the direction of growth, and used continuous-flow isotope ratio mass spectrometry to determine the isotopic compositions of carbon and nitrogen. 
Note that carbon and nitrogen isotopes are expressed as $\delta^{13}$C and $\delta^{15}$N values, i.e. relative to a standard and as ratios of the rare ($^{13}$C, $^{15}$N) to common ($^{12}$C, $^{14}$N) isotope\cite{west2006stable}.

Given the date of stranding (25th March 1891), and estimated baleen growth rates of 13.5cm $y^{-1}$ (see Supplementary Methods), we reconstructed a timeline for $\delta^{13}$C and $\delta^{15}$N fluctuations in the baleen over the seven full years of the whale's life (early 1884 - spring 1891). 
The $\delta^{13}$C and $\delta^{15}$N profiles show two distinct phases in the whale's behaviour. 
Throughout the record, $\delta^{15}$N values record regular cyclical fluctuations, with a mean spacing of 13.5cm, assumed to represent annual periodicity (see Supplementary Methods).  
In behavioural phase one (from record start in early 1884 to spring 1886), we find relatively stable, elevated $\delta^{13}$C values, and relatively low $\delta^{15}$N values (Fig. \ref{fig1}; Figs. S1 and S2). 
In behavioural phase two (summer 1886 to spring 1891) $\delta^{15}$N values are relatively elevated and we see coincident cyclical fluctuations in both $\delta^{13}$C and $\delta^{15}$N values. 
In the last year of life the cyclical pattern is disrupted, with constant low $\delta^{13}$C values for approximately six months in the first half of 1890, before a rapid switch to relatively positive $\delta^{13}$C values in the second half of 1891. 
The final three months of the record show a continued fall in $\delta^{13}$C values (Fig. \ref{fig1}; Figs. S1 and S2).

@@

% figure 1
\begin{figure}
  \centering
  \includegraphics[width = \linewidth]{figures/Figure-1-raw-dC-dN-data.png}
  \caption{Variation in stable isotope values in the NHM blue whale, expressed as $\delta^{13}$C (black circles, left y-axis) and $\delta^{15}$N (grey circles, right y-axis). Samples were taken longitudinally through the baleen plate (n = XX samples from a single baleen plate for both isotopes) of the NHM blue whale exhibits strong annual periodicity and cross-correlation (Figs. S1 and S2) in both isotopes as well as a change in the mean and magnitude of the cycles during 1889 (approximately at sample 30cm). Approximate relationship to years assuming a growth rate of 13.5 cm y$^{-1}$ is shown on the upper x-axis, and year boundaries are indicated by vertical dotted grey lines.}
  \label{fig1}
\end{figure}

% Krill has a very similar isotopic composition to phytoplankton

Relating marine animal tissue isotopic compositions to likely locations is complicated by both a lack of spatial reference values, and because the isotopic composition of phytoplankton at the base of marine food webs varies in both space and time. %line 82
We approach these problems using newly developed, temporally dynamic models of $\delta^{13}$C values in the global ocean\cite{magozzi2017using} (Fig. S3), coupled to an agent-based model of whale movements (see Supplementary Methods; Table S1; Fig. S5) to simulate time series of $\delta^{13}$C values expected under differing movement behaviours and in differing parts of the known geographic range of blue whales. 
We then compare simulated profiles (Figs. \ref{fig2}, \ref{fig3} and Fig. S4) with the measured isotopic profile of the blue whale's baleen (Fig. \ref{fig1}) to identify the most likely movement history.
We assume continued ingestion of new protein throughout the year. 
Other mysticete whales fast during migrations and during winter, relying on stored lipid reserves to meet metabolic demands, but blue whales are too large to stop feeding for long periods; \cite{busquets2017estimating,silva2013north}).
%Can we back this up? Are we more saying that we think they need protein to maintain tissue turnover but could meet energetic demands form stored lipids?

@@

line 112 - 122 para
The last 500 days of the whale's life are difficult to simulate. 
Beginning in the winter of 1889/1890, the observed $\delta^{13}$C profile is relatively depleted, and remains constant for approximately six months, before increasing rapidly in the winter of 1890/1891. 
Depleted $\delta^{13}$C values are found in northern waters that show large temporal fluctuations in $\delta^{13}$Cplk values [fig SX residency plot], and the observed values cannot be simulated purely from movements within the known geographic range.
We infer that the $\delta^{13}$C record in the last period of life reflects the physiological demands of pregnancy, as similar increases in $\delta^{13}$C values have been seen in pregnant South American sea lions\cite{cardona2017temporal}. 
We suggest that the constant, depleted $\delta^{13}$C values reflect use of carbon reserves assimilated in northern latitudes during 6-8 months residency in low latitude waters. 
After six months the whale began feeding in low latitude waters producing a dramatic increase in $\delta^{13}$C values and began a final migration north. 
Blue whales have a 10-12 month gestation period, with calving occurring in sub-tropical waters, and calves are weaned after 6-7 months\cite{handbook}. 
The proposed timescale of pregnancy around early spring 1889, birth in spring 1890, residency in warm waters through the whole of 1890 with feeding resuming in the second half of 1890 and eventual stranding during the return to northern feeding grounds in early 1891, is therefore consistent with the known reproductive ecology of blue whales.

@@

Conventional wisdom suggests that baleen whales migrate each year from feeding grounds near the poles in summer, to overwintering and breeding grounds nearer the equator\cite{corkeron1999baleen,lockyer1981migration}. 
However, a growing body of evidence suggests that not all whales migrate to subtropical waters each year, with some remaining at high latitudes year round\cite{mcdonald2006biogeographic}. %line 127
Our results support this conclusion. 
Interestingly, there is still a resident%line 128
 population of blue whales in the regions indicated by our model, and females still breed in the subtropical waters near the Azores\cite{reilly2008balaenoptera}. 
Whaling was an intense pressure for blue whales during the period we are analysing. 
Before whaling in the North Atlantic began in 1868\cite{reilly2008balaenoptera}, there were 10,000-15,000 blue whales in the region\cite{sigurjonsson1995life}. 
In the early 20th century, fisheries moved outside the area because stocks were so depleted\cite{reilly2008balaenoptera}; during this period over 12,000 blue whales were landed\cite{sigurjonsson1995life}. 
The NHM blue whale thus represents an individual from a species at the brink of extinction.
Our data suggest that the spatial ecology of blue whales in the North Atlantic has remained similar for more than 100 years, despite dramatic changes in anthropogenic pressures. % line 137 and 139

This is the first quantitative evidence of historical individual-level movement behaviours in blue whales. %line 140-141
Our novel simulation modeling removes a long-standing limitation in stable isotope ecology, and can be applied to stable isotope records from any incrementally-grown tissue to estimate most likely individual movement behaviours over multiple years. 
Such techniques can also be applied to extinct species behaviours and used to understand how populations were affected by past pressures such as hunting, and present-day pressures such as global change. 
Finally, our results highlight that museums represent a vast, mostly untapped, source of individual-level spatial ecology information. %line 147-149
